---
title: "clustering-based-demultiplexing"
output: 
  html_document:
    toc: true
    theme: united
# rmarkdown::html_vignette

vignette: >
  %\VignetteIndexEntry{clustering-based-demultiplexing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette provides a brief introduction on how to run the clustering-based demultiplexing workflow for single-cell sample multiplexing experiments. This workflow takes as input tag counts matrix with tags as rows and cell IDs as columns and returns sample-of-origin for each cell. This workflow starts by clustering cells with similar tags expression profiles. Then, for each cluster marker tags are identified and cells are classified depending on marker tags as follows, - If a cluster have at least two markers, cells belong to this cluster are classified as doublets - If a cluster has only one marker, cells belong to this cluster are classified as singlets. - If a cluster has no marker, cells belong to this cluster are classified as negatives


## Loading necessary packages

```{r setup, message=FALSE, warning=FALSE}
library(hashDemux)
library(Seurat)
library(ComplexHeatmap)
library(doParallel)
library(dplyr)
```


## downloading sample tag counts data

```{r, message=FALSE, warning=FALSE}

# downloading tag counts matrix from (Howeitt et. al., 2022)
download.file("https://raw.githubusercontent.com/Oshlack/hashtag-demux-paper/main/data/BAL_data/batch3_c1_hto_counts.csv", "hto_counts.csv")

# downloading ground truth cell assignments
download.file("https://raw.githubusercontent.com/Oshlack/hashtag-demux-paper/main/data/BAL_data/batch3_c1_donors.csv", "ground_truth.csv")

# tags IDs and their corresponding donor IDs
# https://github.com/Oshlack/hashtag-demux-paper/blob/main/analysis/BAL_analysis.Rmd
HTO_donor_list_batch3 <- list("BAL 17" = "BAL Q", 
                              "BAL 18" = "BAL R", 
                              "BAL 19" = "BAL S", 
                              "BAL 20" = "BAL T", 
                              "BAL 21" = "BAL U", 
                              "BAL 22" = "BAL V", 
                              "BAL 23" = "BAL W", 
                              "BAL 24" = "BAL X", 
                              "Doublet" = "Doublet", 
                              "Negative" = "Negative")

ground_truth = read.csv("ground_truth.csv", header = T,row.names = 1)
rownames(ground_truth) = ground_truth$Barcode
table(ground_truth$genetic_donor)

hto_counts = read.csv("hto_counts.csv", header = T, row.names = 1)
rownames(hto_counts) = HTO_donor_list_batch3[rownames(hto_counts)]

```

## creating Seurat object for downsampled data

```{r, message=FALSE, warning=FALSE}
seurat_object = CreateSeuratObject(counts = hto_counts,assay = "HTO", names.delim = "")
seurat_object

seurat_object$ground_truth = ground_truth$genetic_donor

# downsample dataset
seurat_object <- seurat_object[, sample(colnames(seurat_object), size = ncol(seurat_object)/10, replace=F)]
```

## generate UMAP plot

```{r, message=FALSE,  fig.dim = c(7, 5), warning=FALSE}
assay = "HTO"
seurat_object = seurat_object %>% NormalizeData(assay = assay, normalization.method = "CLR",margin = 2) 
seurat_object = seurat_object %>% ScaleData()  
seurat_object = seurat_object %>% RunPCA(features = rownames(seurat_object[[assay]]),npcs=nrow(seurat_object[[assay]]) - 1)
seurat_object = seurat_object %>%  RunUMAP(dims = 1:(nrow(seurat_object[[assay]]) - 1), seed.use = 1)

DimPlot(seurat_object, group.by = "ground_truth")
```

## clustering-based demultiplexing

```{r, message=FALSE, fig.dim = c(10, 6),warning=FALSE}
seurat_object = clustering_based_demux(seurat_object, assay = "HTO")

#table(seurat_object$classification)
seurat_object$clustering_based = seurat_object$sampleBC
seurat_object$clustering_based[seurat_object$classification == "Doublet"] = "Doublet"

DimPlot(seurat_object,group.by = "clustering_based", label = T) + DimPlot(seurat_object,group.by = "ground_truth", label = T)

```

## comparison with HTOdemux

```{r, fig.dim = c(10, 7), message=FALSE, warning=FALSE}
seurat_object = seurat_object %>% NormalizeData(assay = assay, normalization.method = "CLR",margin = 1)  
seurat_object = HTODemux(seurat_object)
seurat_object$HTOdemux = seurat_object$hash.ID

seurat_object$HTOdemux = factor(seurat_object$HTOdemux, levels = sort(levels(seurat_object$HTOdemux)))
#DimPlot(seurat_object, group.by = "HTOdemux") + DimPlot(seurat_object, group.by = "clustering_based")+
#  DimPlot(seurat_object, group.by = "ground_truth")


col_fun = circlize::colorRamp2(breaks = c(0,0.5 , 1),colors = c("lightgray", "yellow" , "red" ) , reverse = F)

mtrx = table(seurat_object$ground_truth, seurat_object$clustering_based) %>% as.matrix()
mtrx2 = apply(mtrx , MARGIN = 1,FUN = function(x) x / sum(x)) %>% t()

clustering_heatmap= Heatmap( mtrx2 , cluster_columns = F, cluster_rows = F,col = col_fun,
           row_title = "Ground Truth", column_title_side = "bottom",
           column_title = "Predicted label (Clustering-based)",
           name = "Proportion of cells",   show_heatmap_legend = F,
           cell_fun = function(j, i, x, y, width, height, fill) 
             {grid.text(sprintf("%d", mtrx[i, j]), x, y, gp = gpar(fontsize = 10))},
           row_names_side = "left", column_names_side = "top" )

mtrx = table(seurat_object$ground_truth, seurat_object$HTOdemux) %>% as.matrix()
mtrx2 = apply(mtrx , MARGIN = 1,FUN = function(x) x / sum(x)) %>% t()

htodemux_heatmap= Heatmap( mtrx2 , cluster_columns = F, cluster_rows = F,col = col_fun,
           row_title = "Ground Truth", column_title_side = "bottom",
           column_title = "Predicted label (HTOdemux)",
           name = "Proportion of cells2",  show_heatmap_legend = F,
           cell_fun = function(j, i, x, y, width, height, fill) 
             {grid.text(sprintf("%d", mtrx[i, j]), x, y, gp = gpar(fontsize = 10))},
           row_names_side = "left", column_names_side = "top" )

htodemux_heatmap + clustering_heatmap
```

## Session Info

```{r}
sessionInfo()
```
